---
title: "The SubnationalCRVS R Package"
author: "Jeremy Roth"
date: ''
output:
  pdf_document:
    number_sections: yes
    toc: yes
bibliography: bibliography.bib
header-includes:
 \usepackage{float}
---

```{r "clear", include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
The `SubnationalCRVS` package, which is still under active development, is hosted on my GitHub page (`www.github.com/jroth-unfpa/SubnationalCRVS`) rather than on CRAN. As a result, `SubnationalCRVS` must be installed with the `install.github()` function from the `devtools` package instead of the usual `install.packages()` function. The required dependency `DemoTools` [@DemoTools] is also hosted on GitHub instead of CRAN and thus must also be installed with `install.github()`.

```{r "install_package", eval=FALSE}
library(devtools)
install_github("timriffe/DemoTools") # install the DemoTools dependency 
install_github("jroth-unfpa/SubnationalCRVS") # install the SubnationalCRVS package
```
Now we can load `SubnationalCRVS`, specify the name of the local folder in which we will save the plots produced by the package, and create that local folder if it does not already exist.
```{r "load_package", message=FALSE, warning=FALSE}
library(SubnationalCRVS)
my_plots_dir <- "Plots/" # local folder where the plots should be saved
dir.create(my_plots_dir) # create the folder if it does not already exist
```

For this demonstration, we will also load the `dplyr` package to customize the display of some of the tables returned by `SunbationalCRVS`.
```{r "load_dplyr", message=FALSE, warning=FALSE}
library(dplyr)
```

# Datasets Included with the Package: Tabulations from Ecuador
The `SubnationalCRVS` package comes with tabulations of population and registered deaths in Ecuador -- disaggregated by age, sex, and province -- created from publicly available datasets from Ecuador [@Peralta:2019; @INEC:2010] based on the Ecuador's 2001 Census, 2010 Census, and annual counts of registered deaths from 2001 through 2010. The structure of the included example datasets is inspired by the data requirements of the `ddm()` function from the `DDM` package [@DDM].

The `ecuador_single_year_ages_combined` dataset reports the estimated populations in Ecuador  by single-year ages (0, 1, 2, ... in the `age` column) from both the 2001 Census (`pop1` column) and the 2010 Census (`pop2` column), separately for males and females (`m` and `f` in the `sex` column) and province (in the `province` column).

```{r "ecuador_single_year_ages", echo=TRUE}
head(ecuador_single_year_ages)
```

The `ecuador_single_year_ages_combined` dataset appends rows to `ecuador_single_year_ages_combined` that  report the sex- and single-year-age- disaggregated population estimates for the entire country (as opposed to a single province).

```{r "ecuador_single_year_ages_combined", echo=TRUE}
ecuador_single_year_ages_combined %>% filter(province_name == "National") %>%
                                    head()
```

The `ecuador_five_year_ages` and `ecuador_five_year_ages_combined` datasets have the same variables as `ecuador_single_year_ages_combined` and `ecuador_single_year_ages_combined` with two exceptions: (1) the `age` variable now represents five-year age groups (in the `age` column, with 0-4 coded as 0, 5-9 coded as 5, 10-14 coded as 10, etc.) instead of single-year ages; and (2) there is an additional column called `deaths` that reports the registered deaths collected between 2001 and 2010.

```{r "ecuador_five_year_ages", echo=TRUE}
head(ecuador_five_year_ages)
```

```{r "ecuador_five_year_ages_combined", echo=TRUE}
head(ecuador_five_year_ages_combined)
ecuador_five_year_ages_combined %>% filter(province_name == "National") %>%
                                    head()
```

# Example of Subnational Analysis: Provinces in Ecuador
In this section, we describe the key functions of `SubnationalCRVS` in the context of visualizing outputs from a demographic data quality assessment (DDQA) and estimates of death registration completeness [@DDM] within provinces of Ecuador. Later in this tutorial, we also show how `SubnationalCRVS` provides corresponding visualizations for national-level results.

## Demographic Data Quality Assessment (DDQA)
### Sex ratios: PlotSexRatios()
A key step in this demographic data quality assessment (DDQA) process is to use the `PlotSexRatios()` function to compute and plot sex ratios within each combination of province, sex, and single-year age for the 2001 and 2010 data stored in `ecuador_single_year_ages_combined`.

To use `PlotSexRatios()`, we are required to provide our tabulated data frame in the `data` argument and a few additional required arguments that describe the variable names and values in `data`. These arguments and the expected format of the specified dataset are motivated by the data structure enforced in the `DDM` package [@DDM].

- `name.disaggregations` is the name of variable representing the subnational disaggregation (apart from sex, which is required,) in `data`. Here we specify `name.disaggregations="province_name"`.

- `name.sex` is the name of variable representing sex. Here we specify `name.sex="sex"`

- `name.age` is the name of variable representing age. Here we specify `name.age="age"`

- `name.date1` is the name of variable that provides the date of the earlier of the two time periods represented in `data. Here we specify `name.date1="date1"` (the value of this variable, `"2001-11-15"` is the date of Ecuador's 2001 Census)

- `name.date2` is the name of variable that provides the date of the earlier of the two time periods represented in `data`. Here we specify `name.date2="date2"` (the value of this variable, `"2010-11-28"` is the date of Ecuador's 2010 Census)

- `name.population.year1` is the name of variable representing the population in the earlier of the two time periods represented in the dataset. Here we specify `name.population.year2="pop1"`

- `name.population.year2` is the name of variable representing the population in the earlier of the two time periods represented in the dataset. Here we specify `name.population.year2="pop2"`

- `name.male` is the name of value of the `name.sex` variable that represents males. Here we specify `name.males="m"`

- `name.female` is the name of value of the `name.sex` variable that represents females. Here we specify `name.females="f"`

In addition, we specify two optional arguments, `plots.dir` and `name.national`. We set `plots.dir=my_plots_dir` so that the plots will save in the local folder `Plots`; if the `plots.dir` argument is omitted, the plots will be saved in the same working directory of the R script. We also set `name.national="National"` to reflect the fact that national-level datasets are provided in rows where the `name.disaggregations` variable has the value `"National"`.  Specifying `name.national` produces visualizations catered specifically to national-level analysis (instead of accommodated further disaggregation) that will be presented later in the tutorial. If `name.national` is not provided, those national-level visualizations are simply not produced.

```{r "compute_sex_ratios", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
c_s <- ComputeSexRatios(data=ecuador_single_year_ages_combined,
                   name.disaggregations="province_name",
                   name.males="m",
                   name.females="f",
                   name.age="age",
                   name.sex="sex",
                   name.date1="date1",
                   name.date2="date2",
                   name.population.year1="pop1",
                   name.population.year2="pop2")
```

```{r "plot_sex_ratios", message=FALSE, warning=FALSE}
s <- PlotSexRatios(data=ecuador_single_year_ages_combined,
                   name.disaggregations="province_name",
                   name.males="m",
                   name.females="f",
                   name.age="age",
                   name.sex="sex",
                   name.date1="date1",
                   name.date2="date2",
                   name.population.year1="pop1",
                   name.population.year2="pop2",
                   name.national="National",
                   label.subnational.level="Province",
                   plots.dir=my_plots_dir)
```
The plots of sex ratios are saved in the `Plots/` sub-folder we specified with the argument `plots.dir=my_plots_dir`; `plots.dir` is an optional argument and, if we do not specify a value for it, the plots will be saved the working directory. We also specified `label.subnational.level="Province"` so that the disaggregations are labeled `Province` instead of the less clear `province_name`.

The sex ratios for all levels of subnational disaggregation are overlaid in the following "combined" plots separately for each data year.
```{r "plot_sex_ratio_combined", echo=TRUE, fig.cap="Sex ratios in Ecuador by province, combined plot", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "sex_ratios_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

Additionally, the sex ratios are plotted in separate figures for each level of subnational disaggregation in the following "disaggregated" plots.

```{r "plot_sex_ratio_disaggregated", echo=TRUE, fig.cap="Sex ratios in Ecuador by province, disaggregated plots (only first page shown)", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "sex_ratios_by_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
``` 

### View sex ratios in table
The object returned by `PlotSexRatios()` is a table that shows us the sex ratios for each combination of province, sex, and single-year age in the `sex_ratio_1` column (for the 2001 Census) and the `sex_ratio_2` column (for the 2010 Census).

```{r ¨sex_ratios_table"}
s %>% select(province_name, age, pop1, pop2, sex_ratio_1, sex_ratio_2) %>%
      head()
```

### Age ratios: PlotAgeRatios()
Another step in our demographic data quality assessment is using the `PlotAgeRatios()` function in the `SubnationalCRVS` package to compute and plot age ratios within each combination of province and sex for the 2001 and 2010 data stored in `ecuador_five_year_ages_combined`. The arguments we provide to `PlotAgeRatios()` are actually identical to those we specified for `PlotSexRatios()`, except now we are using the tabulation with five-year age groups (`ecuador_five_year_ages_combined`) instead of the tabulation with single-year ages.

```{r "age_ratios", message=FALSE, warning=FALSE, eval=TRUE}
a <- PlotAgeRatios(data=ecuador_five_year_ages_combined,
              name.disaggregations="province_name",
              name.males="m",
              name.females="f",
              name.age="age",
              name.sex="sex",
              name.date1="date1",
              name.date2="date2",
              name.population.year1="pop1",
              name.population.year2="pop2",
              label.subnational.level="Province",
              name.national="National",
              plots.dir=my_plots_dir)
```


### View age ratios in combined plot
The following "combined" plots, saved in the `Plots/` folder, show the age ratios for all levels of subnational disaggregation, separately for males and females in each data year.
```{r "plot_age_ratios_combined", echo=TRUE, fig.cap="Age ratios in Ecuador by province, combined plot", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_ratios_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

In addition, `PlotAgeRatios()` also creates the following "disaggregated" plots, saved in the `Plots/` folder, where the age ratios for each level of disaggregation are shown in separate plots, with different sexes and data years overlaid within each plot.

```{r "plot_age_ratios_disaggreagted", echo=TRUE, fig.cap="Age ratios in Ecuador by province, disaggregated plots (only first page shown) ", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_ratios_by_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

Just as the `PlotSexRatios()` function returns a table of disaggregated sex ratios, `PlotAgeRatios()` returns a table of disaggregated age ratios.
```{r "age_ratios_table"}
a %>% select(province_name, age, pop1, pop2, age_ratio_1, age_ratio_2) %>%
      head()
```

### Potential age heaping: PlotPotentialAgeHeaping()
To give us a sense of whether "age-heaping" is occurring within the levels of disaggregation present in our `ecuador_single_year_ages_combined` dataset, we turn to the `PlotPotentialAgeHeaping()` function and actually provide the same arguments we used in the `PlotSexRatios()` function.

```{r "age_heaping", message=FALSE, warning=FALSE, eval=TRUE}
PlotPotentialAgeHeaping(data=ecuador_single_year_ages_combined,
            name.disaggregations="province_name",
              name.males="m",
              name.females="f",
              name.age="age",
              name.sex="sex",
              name.date1="date1",
              name.date2="date2",
              name.population.year1="pop1",
              name.population.year2="pop2",
              name.national="National",
              label.subnational.level="Province",
              plots.dir=my_plots_dir)
```

### View potential age heaping in combined plot
The following "combined" plots show us estimated population counts by single-year ages with different provinces represented with different overlaid colors, and separate plots for each sex and data year.
  
```{r "plot_age_heaping_combined", echo=TRUE, fig.cap="Population counts in Ecuador by single-year age, combined plot", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "potential_age_heaping_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

`PlotPotentialAgeHeaping()` present separate plots of population counts for each province in Ecuador, with different sexes and data years overlaid within each plot.
```{r "plot_age_heaping_disaggregated", echo=TRUE, fig.cap="Population counts in Ecuador by single-year age, disaggregated plots", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "potential_age_heaping_by_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

### Age heaping indices: PlotAgeHeapingScores() 
As a more concise summary of potential age-heaping suggested by the visualizations from `PlotPotentialAgeHeaping`, we now use the `PlotAgeHeapingScores` function with the same arguments we provided to the `PlotAgeRatios` function.

```{r "age_heaping_scores", message=FALSE, warning=FALSE, eval=TRUE}
ageheaping <- PlotAgeHeapingScores(data=ecuador_single_year_ages_combined,
                                      name.disaggregations="province_name",
                                      name.males="m",
                                      name.females="f",
                                      name.age="age",
                                      name.sex="sex",
                                      name.date1="date1",
                                      name.date2="date2",
                                      name.population.year1="pop1",
                                      name.population.year2="pop2",
                                      name.national="National",
                                      label.subnational.level="Province",
                                      plots.dir=my_plots_dir)
```

One set of plots returned by `PlotAgeHeapingScores()` shows the values of three age-heaping indices within each combination of province, sex, and data year: Roughness, Whipple, and Myers. The indices are computed with the `check_heaping_roughness()`, `check_heaping_Whipple()`, and `check_heaping_myers()` functions, respectively, from the `DemoTools` package [@DemoTools].

```{r "plot_age_heaping_indices", echo=TRUE, fig.cap="Roughness, Whipple, and Myers indices in Ecuador by province (only first page shown)", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_heaping_scores_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

Further, `PlotAgeHeapingScores()` also creates plots showing the Noumbissi index in each province computed for each terminal digit from 0-9, using the `check_heaping_Noumbissi()` function from the `DemoTools` package [@DemoTools].

```{r "plot_Noumbissi", echo=TRUE, fig.cap="Noumbissi indices in Ecuador by province (only first page shown)", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "Noumbissi_by_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

We can also view the age-heaping indices in the table returned by `PlotAgeHeapingScores()`

```{r "age_heaping_indices_table"}
head(ageheaping)
```

## DDM Estimation of Death Registration Completeness
The structure of the `ecuador_five_year_ages` dataset is inspired by the requirements for the `ddm()` function from the `DDM` package [@DDM], which uses established Death Distribution Methods (DDM) to estimate death registration completeness for adults (aged 15+) between two consecutive Censuses. Essentially, the `SubnationalCRVS` package offers its `EstimateDDM()` function only as a convenient wrapper to `DDM::ddm()` to perform DDM estimation of death registration completeness using the same pipeline used to perform the DDQA.

### Compute DDM estimates: EstimateDDM()
```{r "estimate_DDM", message=TRUE, warning=FALSE, eval=TRUE}
ddm_results <- EstimateDDM(data=ecuador_five_year_ages_combined, 
            name.disaggregations="province_name",
            name.age="age",
            name.sex="sex",
            name.males="m",
            name.females="f",
            name.date1="date1",
            name.date2="date2",
            name.population.year1="pop1",
            name.population.year2="pop2",
            name.national="National",
            name.deaths="deaths",
            deaths.summed=TRUE)
```

We note `EstimateDDM()` uses exactly the same arguments as `PlotAgeRatios()`, for example, from the DDQA with two additional required arguments:

- `name.deaths`, which provides the name of the variable representing the count of registered deaths between the two dates represented in `name.date1` and `name.date2`

- `deaths.summed`, which should be set to TRUE when the `name.deaths` variable represents the total number of registered deaths `name.date1` and `name.date2` and set to FALSE when the `name.deaths` variable represents the average number of registered deaths between the two dates.

### Plot DDM estimates: PlotDDM()
We can plot the estimated adult death registration completeness (using the "hybrid" GGB-SEG method) with the `PlotDDM` function:
```{r "plot_DDM", message=FALSE, warning=FALSE, eval=TRUE}
PlotDDM(ddm_results=ddm_results,
        label.subnational.level="Province",
        plots.dir=my_plots_dir)
```

```{r "DDM_point_estimates_in_plot", echo=TRUE, fig.cap="Point estimates of death registration completeness in Ecuador from 2001-2010, using the GGB-SEG method", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "ggbseg_point_estimates_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

The `EstimateDDM` function also returns a list, in which the `ddm_estimates` object contains the estimated death registration completeness with the GGB-SEG approach 
```{r "DDM_point_estimates_table", message=TRUE, warning=FALSE, eval=TRUE}
head(ddm_results$ggbseg_estimates)
```

Additionally, `PlotDDM` presents a visualization

of the point estimates of completeness and the corresponding root mean squared errors (RMSEs) for each of the permitted values of the age range that is selected `DDM::ddm()` as part of its fitting procedure.

```{r "DDM_sensitivity_estimates_in_plot", echo=TRUE, fig.cap="Sensitivity of point estimates of death registration completeness in Ecuador from 2001-2010 to choice of age-range parameter in the GGB-SEG method", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "ggbseg_sensitivity_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

```{r "DDM_RMSE_estimates_in_plot", echo=TRUE, fig.cap="Sensitivity of point estimates of death registration completeness in Ecuador from 2001-2010 to choice of age-range parameter in the GGB-SEG method", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "ggbseg_RMSE_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

The sensitivity estimates are also returned in table by `EstimateDDM`, in the `sensitivity_ddm_estimates` element of its list.
```{r "DDM_sensitivity_estimates_table"}
head(ddm_results$sensitivity_ggbseg_estimates, n=5)
```

# Example of National Analysis: Ecuador
As discussed, specifying the argument `name.national="National"` during the previous DDQA and estimation of death registration completeness permitted `SubnationalCRVS` to save plots specifically designed to display national-level (as opposed to subnational-level) visualizations. These national-level plots are can also be produced by specifying a national-level-only dataset, for example `ecuador_five_year_ages_national` (shown below), as long as the corresponding `name.national` argument is also specified (e.g. `name.national="National"` here)

```{r "show_five_years_national"}
head(ecuador_five_year_ages_national)
tail(ecuador_five_year_ages_national)
```

## Plot of sex ratios
```{r "plot_sex_ratio_national", echo=TRUE, fig.cap="Sex ratios in Ecuador", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "sex_ratios_national_",
                                   Sys.Date(),
                                   ".pdf"))
```

## Plot of age ratios
```{r "plot_age_ratios_national", echo=TRUE, fig.cap="Age ratios in Ecuador", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_ratios_national_",
                                   Sys.Date(),
                                   ".pdf"))

```

## Plots related to age heaping
### Plot of potential age heaping
```{r "plot_age_heaping_combined_national", echo=TRUE, fig.cap="Population counts in Ecuador by single-year age", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "potential_age_heaping_National_",
                                   Sys.Date(),
                                   ".pdf"))

```
### Plot of age heaping indices
```{r "plot_age_heaping_indices_national", echo=TRUE, fig.cap="Roughness, Whipple, and Myers indices in Ecuador", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_heaping_scores_National_",
                                   Sys.Date(),
                                   ".pdf"))
```

### Plot of Noumbissi indices
```{r "plot_Noumbissi_national", echo=TRUE, fig.cap="Noumbissi indices in Ecuador", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "Noumbissi_National_",
                                   Sys.Date(),
                                   ".pdf"))
```

## Plot of DDM estimates
```{r "DDM_sensitivity_estimates_in_plot_national", echo=TRUE, fig.cap="Sensitivity of point estimates of death registration completeness in Ecuador from 2001-2010 to choice of age-range parameter in the GGB-SEG method", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "ggbseg_sensitivity_National_",
                                   Sys.Date(),
                                   ".pdf"))
```
```{r "DDM_RMSE_estimates_in_plot_national", echo=TRUE, fig.cap="Sensitivity of point estimates of death registration completeness in Ecuador from 2001-2010 to choice of age-range parameter in the GGB-SEG method", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "ggbseg_RMSE_National_",
                                   Sys.Date(),
                                   ".pdf"))
```

# References



