---
title: "SubnationalCRVS R Package: Demo"
author: "Jeremy Roth"
date: ''
output:
  pdf_document:
    number_sections: yes
    toc: yes
header-includes:
 \usepackage{float}
---

```{r "clear", include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
Since the `SubnationalCRVS` package is still under active development, it is hosted on my GitHub page (`www.github.com/jroth-unfpa/SubnationalCRVS`) rather than on CRAN. As a result, `SubnationalCRVS` cannot be installed with the usual `install.packages()` function; instead, we install `SubnationalCRVS` with the `install.github()` function from the `devtools` package. The dependency `DemoTools` is also not hosted on CRAN and must be installed with `install.github()`.
```{r "install_package", eval=FALSE}
library(devtools)
install_github("timriffe/DemoTools") # installing the DemoTools dependency (not on CRAN)
install_github("jroth-unfpa/SubnationalCRVS") # installing the SubnationalCRVS package itself
```
Now we can load `SubnationalCRVS`, specify the name of the local folder in which we will save the plots, and create that local folder if it does not already exist.
```{r "load_package", message=FALSE, warning=FALSE}
library(SubnationalCRVS)
my_plots_dir <- "Plots/" # local folder where the plots should be saved
dir.create(my_plots_dir) # create the folder if it does not already exist
```

We will also load the `dplyr` package to customize the display of some of the tables returned by `SunbationalCRVS`.
```{r "load_dplyr", message=FALSE, warning=FALSE}
library(dplyr)
```

# View the Example Datasets Included with the SubnationalCRVS Package
The `SubnationalCRVS` package comes with two tabulations -- disaggregated by age, sex, and province -- created from publicly available datasets from Ecuador (LINKS), based on the country's 2001 Census, 2010 Census, and annual counts of registered deaths from 2001 through 2010. 

```{r "ecuador_single_year_ages", echo=TRUE}
head(ecuador_single_year_ages)
```
The `ecuador_single_year_ages` dataset reports the estimated populations in Ecuador  by single-year ages (0, 1, 2, ... in the `age` column) from both the 2001 Census (`pop1` column) and the 2010 Census (`pop2` column), separately for males and females (`m` and `f` in the `sex` column) and province (full name in the `province` column and abbreviated name in the `province_name_short` column).


```{r "ecuador_five_year_ages", echo=TRUE}
head(ecuador_five_year_ages)
```
The `ecuador_five_year_ages` dataset has the same variables as `ecuador_single_year_ages` with two exceptions: (1) the `age` variable now represents five-year age groups (in the `age` column, with 0-4 coded as 0, 5-9 coded as 5, 10-14 coded as 10, etc.) instead of single-year ages; and (2) there is an additional column called `deaths` that reports the registered deaths collected between 2001 and 2010.

# Conduct Demographic Data Quality Assessment (DDQA)

## Sex ratio: PlotSexRatios()
One step in the demographic data quality assessment (DDQA) process is to use the `PlotSexRatios()` function in the `SubnationalCRVS` package to compute and plot sex ratios within each combination of province, sex, and single-year age for the 2001 and 2010 data stored in `ecuador_single_year_ages`.

We must provide our tabulated data frame in the `data` argument of `PlotSexRatios()` and, additionally, we need to provide a few additional arguments to `PlotSexRations()` to describe the variable names and values in the dataset:

- `name.disaggregations` is the name of variable representing the subnational disaggregation (apart from sex) in `data`. Here we specify `name.disaggregations="province_name"`.

- `name.sex` is the name of variable representing sex. Here we specify `name.sex="sex"`

- `name.age` is the name of variable representing age. Here we specify `name.age="age"`

- `name.date1` is the name of variable that provides the data of the earlier of the two time periods. Here we specify `name.date1="date1"` (the value of this variable, `"2001-11-15"` is the date of Ecuador's 2001 Census)

- `name.date2` is the name of variable that provides the data of the earlier of the two time periods. Here we specify `name.date1="date1"` (the value of this variable, `"2010-11-28"` is the date of Ecuador's 2010 Census)

- `name.population.year1` is the name of variable representing the population in the earlier of the two time periods represented in the dataset. Here we specify `name.population.year2="pop1"`

- `name.population.year2` is the name of variable representing the population in the earlier of the two time periods represented in the dataset. Here we specify `name.population.year2="pop2"`

- `name.male` is the name of value of the `name.sex` variable that represents males

- `name.female` is the name of value of the `name.sex` variable that represents females

In addition, we specify the option argument `plots.dir=my_plots_dir` so that the plots will save in a local folder `Plots`; the default behavior would be to save the plots in the working directory of the R script.

```{r "sex_ratios", message=FALSE, warning=FALSE}
c_s <- ComputeSexRatios(data=ecuador_single_year_ages,
                   name.disaggregations="province_name",
                   name.males="m",
                   name.females="f",
                   name.age="age",
                   name.sex="sex",
                   name.date1="date1",
                   name.date2="date2",
                   name.population.year1="pop1",
                   name.population.year2="pop2")

s <- PlotSexRatios(data=ecuador_single_year_ages,
                   name.disaggregations="province_name",
                   name.males="m",
                   name.females="f",
                   name.age="age",
                   name.sex="sex",
                   name.date1="date1",
                   name.date2="date2",
                   name.population.year1="pop1",
                   name.population.year2="pop2",
                   label.subnational.level="Province",
                   plots.dir=my_plots_dir)
```
The plots of sex ratios are saved in the `Plots/` sub-folder we specified with the argument `plots.dir=my_plots_dir`; `plots.dir` is an optional argument and, if we do not specify a value for it, the plots will be saved the working directory. We also specified `label.subnational.level="Province"` so that the disaggregations are labeled `Province` instead of the less clear `province_name`.

### View sex ratios in combined plot 
The sex ratios for all levels of subnational disaggregation are overlaid in the following "combined" plots separately for each data year.
```{r "plot_sex_ratio_combined", echo=TRUE, fig.cap="Sex ratios in Ecuador by province, combined plot", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "sex_ratios_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

### View sex ratios in disaggregated plots
Additionally, the sex ratios are plotted in separate figures for each level of subnational disaggregation in the following "disaggregated" plots
```{r "plot_sex_ratio_disaggregated", echo=TRUE, fig.cap="Sex ratios in Ecuador by province, disaggregated plots", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "sex_ratios_by_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
``` 

### View sex ratios in table
The object returned by `PlotSexRatios()` is a table that shows us the sex ratios for each combination of province, sex, and single-year age in the `sex_ratio_1` column (for the 2001 Census) and the `sex_ratio_2` column (for the 2010 Census).

```{r ¨sex_ratios_table"}
s %>% select(province_name, age, pop1, pop2, sex_ratio_1, sex_ratio_2) %>%
      head()
```

## Age ratios: PlotAgeRatios()
Another step in our demographic data quality assessment is using the `PlotAgeRatios()` function in the `SubnationalCRVS` package to compute and plot age ratios within each combination of province and sex for the 2001 and 2010 data stored in `ecuador_five_year_ages`.
```{r "age_ratios", message=FALSE, warning=FALSE, eval=TRUE}
a <- PlotAgeRatios(data=ecuador_five_year_ages,
              name.disaggregations="province_name",
              name.males="m",
              name.females="f",
              name.age="age",
              name.sex="sex",
              name.date1="date1",
              name.date2="date2",
              name.population.year1="pop1",
              name.population.year2="pop2",
              label.subnational.level="Province",
              plots.dir="Plots/")
```
The arguments we provided to `PlotAgeRatios()` are actually identical to those we specified for `PlotSexRatios()`, except now we are using the tabulation with five-year age groups (`ecuador_five_year_ages`) instead of the tabulation with single-year ages.

### View age ratios in combined plot
The following "combined" plots, saved in the `Plots/` folder, show the age ratios for all levels of subnational disaggregation, separately for males and females in each data year.
```{r "plot_age_ratios_combined", echo=TRUE, fig.cap="Age ratios in Ecuador by province, combined plot", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_ratios_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

### View age ratios in disaggregated plots
In addition, `PlotAgeRatios()` also creates the following "disaggregated" plots, saved in the `Plots/` folder, where the age ratios for each level of disaggregation are shown in separate plots, with different sexes and data years overlaid within each plot.

```{r "plot_age_ratios_disaggreagted", echo=TRUE, fig.cap="Age ratios in Ecuador by province, disaggregated plots", fig.align="center", fig.pos="H", out.width = "60%"}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_ratios_by_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

### View age ratios in table
Just as the `PlotSexRatios()` function returns a table of disaggregated sex ratios, `PlotAgeRatios()` returns a table of disaggregated age ratios.
```{r "age_ratios_table"}
a %>% select(province_name, age, pop1, pop2, age_ratio_1, age_ratio_2) %>%
      head()
```

## Potential age heaping: PlotPotentialAgeHeaping()
To give us a sense of whether "age-heaping" is occurring within the levels of disaggregation present in our `ecuador_single_year_ages` dataset, we turn to the `PlotPotentialAgeHeaping()` function and actually provide the same arguments we used in the `PlotSexRatios()` function.

```{r "age_heaping", message=FALSE, warning=FALSE, eval=TRUE}
PlotPotentialAgeHeaping(data=ecuador_single_year_ages,
            name.disaggregations="province_name",
              name.males="m",
              name.females="f",
              name.age="age",
              name.sex="sex",
              name.date1="date1",
              name.date2="date2",
              name.population.year1="pop1",
              name.population.year2="pop2",
              label.subnational.level="Province",
              plots.dir="Plots/")
```

### View potential age heaping in combined plot
The following "combined" plots show us estimated population counts by single-year ages with different provinces represented with different overlaid colors, and separate plots for each sex and data year.
```{r "plot_age_heaping_combined", echo=TRUE, fig.cap="Population counts in Ecuador by single-year age, combined plot", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "potential_age_heaping_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

### View potential age heaping in disaggregated plots
`PlotPotentialAgeHeaping()` present separate plots of population counts for each province in Ecuador, with different sexes and data years overlaid within each plot.
```{r "plot_age_heaping_disaggregated", echo=TRUE, fig.cap="Population counts in Ecuador by single-year age, disaggregated plots", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "potential_age_heaping_by_province_name_",
                                   Sys.Date(),
                                   ".pdf"))

```

## Age heaping indices: PlotAgeHeapingScores() 
As a more concise summary of potential age-heaping suggested by the visualizations from `PlotPotentialAgeHeaping`, we now use the `PlotAgeHeapingScores` function with the same arguments we provided to the `PlotAgeRatios` function.

```{r "age_heaping_scores", message=FALSE, warning=FALSE, eval=TRUE}
ageheaping <- PlotAgeHeapingScores(data=ecuador_single_year_ages,
                                      name.disaggregations="province_name",
                                      name.males="m",
                                      name.females="f",
                                      name.age="age",
                                      name.sex="sex",
                                      name.date1="date1",
                                      name.date2="date2",
                                      name.population.year1="pop1",
                                      name.population.year2="pop2",
                                      label.subnational.levels="Province",
                                      plots.dir="Plots/")
```

### View age heaping indices in plots
`PlotAgeHeapingScores` plots the values of three age-heaping indices within each combination of province, sex, and data year: Roughness, Whipple, and Myers. The indices are computed with the `DemoTools::check_heaping_roughness`, `DemoTools::check_heaping_Whipple`, and `DemoTools::check_heaping_myers` functions, respectively, from the `DemoTools` package.

```{r "plot_age_heaping_indices", echo=TRUE, fig.cap="Age heaping indices in Ecuador by province", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "age_heaping_scores_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```

We can also view the age-heaping indices in the table returned by `PlotAgeHeapingScores`

### View age heaping indices in table
```{r "age_heaping_indices_table"}
head(ageheaping)
```

### View Noumbissi age-heaping indices: ComputeAgeHeapingScores()
We can also use the `ComputeAgeHeapingScores` function -- which is called within `PlotAgeHeapingScores` -- with the optional argument `Noumbissi.display=TRUE` to view the Noumbissi index computed for single-year ages ending with 0, 1, 2,..., 9.

```{r "age_heaping_Noumbissi", message=FALSE, warning=FALSE, eval=TRUE}
ageheaping_with_Noumbissi <- ComputeAgeHeapingScores(data=ecuador_single_year_ages,
                                      name.disaggregations="province_name",
                                      name.males="m",
                                      name.females="f",
                                      name.age="age",
                                      name.sex="sex",
                                      name.date1="date1",
                                      name.date2="date2",
                                      name.population.year1="pop1",
                                      name.population.year2="pop2",
                                      Noumbissi.display=TRUE)
head(ageheaping_with_Noumbissi)
```

# DDM Estimation of Death Registration Completeness
The structure of the `ecuador_five_year_ages` dataset is inspired by the requirements for the `DDM::ddm` function from the `DDM` package, which uses established Death Distribution Methods (DDM) to estimate death registration completeness between two consecutive Censuses. As a result, the `SubnationalCRVS` package offers the `EstimateDDM` function as a simple wrapper to `DDM::ddm` to perform DDM estimation of death registration completeness.

## Compute DDM estimates: EstimateDDM()
```{r "estimate_DDM", message=TRUE, warning=FALSE, eval=TRUE}
ddm_results <- EstimateDDM(data=ecuador_five_year_ages, 
            name.disaggregations="province_name",
            name.age="age",
            name.sex="sex",
            name.males="m",
            name.females="f",
            name.date1="date1",
            name.date2="date2",
            name.population.year1="pop1",
            name.population.year2="pop2",
            name.deaths="deaths",
            deaths.summed=TRUE)
```

We called `EstimateDDM` above using the same arguments as we used with `PlotAgeRatios`, with two additions: 

- `name.deaths` provides the name of the variable represented the count of registered deaths between the two dates represented in `name.date1` and `name.date2`

- `deaths.summed`, which should be set to TRUE when the `name.deaths` variable represents the total number of registered deaths `name.date1` and `name.date2` and set to FALSE when the `name.deaths` variable represents the average number of registered deaths between the two dates.

## Plot DDM estimates: PlotDDM()
We can plot the estimated death registration completeness (using the "hybrid" GGB-SEG method) with the `PlotDDM` function:
```{r "plot_DDM", message=FALSE, warning=FALSE, eval=TRUE}
PlotDDM(ddm_results=ddm_results,
        label.completeness="Estimated Completeness (%)",
        label.subnational.levels="Province",
        plots.dir="Plots/")
```

### View estimates of death registration completeness in plot
```{r "DDM_point_estimates_in_plot", echo=TRUE, fig.cap="Point estimates of death registration completeness in Ecuador from 2001-2010, using the GGB-SEG method", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "ddm_point_estimates_combined_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```


### View DDM point estimates in table 
The `EstimateDDM` function also returns a list, in which the `ddm_estimates` object contains the estimated death registration completeness with the GGB-SEG approach 
```{r "DDM_point_estimates_table", message=TRUE, warning=FALSE, eval=TRUE}
head(ddm_results$ggbseg_estimates)
```

### View sensitivity of DDM point estimates in plot
Additionally, `PlotDDM` presents a visualization of the GGB-SEG estimates of death-registration completeness to the all permitted values of the age range (i.e. sensitivity) that is selected `DDM::ddm()` as part of the fitting procedure underlying its estimation.

```{r "DDM_sensitivity_estimates_in_plot", echo=TRUE, fig.cap="Sensitivity of point estimates of death registration completeness in Ecuador from 2001-2010 to choice of age-range parameter in the GGB-SEG method", fig.align="center", fig.pos="H", out.width = "60%", eval=TRUE}
knitr::include_graphics(path=paste0(my_plots_dir,
                                   "ddm_sensitivity_province_name_",
                                   Sys.Date(),
                                   ".pdf"))
```


### View age-range sensitivity of DDM point estimates in table 
The sensitivity estimates are also returned in table by `EstimateDDM`, in the `sensitivity_ddm_estimates` element of its list.
```{r "DDM_sensitivity_estimates_table"}
head(ddm_results$sensitivity_ggbseg_estimates, n=5)
```



